Shapes analysis for time series.

### 2024-05-22 NEURIPS24/paper.tex
update

  In the next section, the time series are going to be represented by diffeomorphism parameters $(\alpha_j)_{j\in[N]}$.
  That's why LDDMM is chosen since it offers a parametrization for diffemorphisms which is sparse and interpretable, two features particularly relevant in the biomedical context.

The basic problem that we consider in this section is the following. Given a set of targets $\mathbf{y}=(y_i)_{i\in[T_2]}$ in $\Rset^{d'}$\footnote{Note that we denote by $d'\in\nset$ the ambient space}, a set of starting points $\mathbf{x}=(x_{i})_{i\in[T_1]}$ in $\Rset^{d'}$, we aim to find a diffeomorphism $\phi$ such that the finite set of points $\mathbf{y}$ is similar in a certain sense to the set of finite sets of transformed points $\phi \cdot \mathbf{x} =(\phi(x_i))_{i\in[T_1]} $.
 The function $\phi$ is occasionally referred to as a \textit{deformation}. In general, these sets $\mathbf{x},\mathbf{y}$ are meshes of continous objects, e.g surfaces, curves, images and so on.

@@

%\paragraph*{Regularizing the problem to derive uniqueness.}
It has been proposed in \cite{miller2006geodesic} to interpret $\msv$ as a tangent space relative to the group of diffeomorphisms $\msh=\{ \phi^v:\eqsp v\in \rml^2([0,1],\msv)\}$.
Following this geometric point of view, geodesics can be constructed on $\msh$ by using the following squared norm 
 \begin{equation}
  \label{eq:geodesics_original}
    \mathscr{R}^2: g\in \msh\mapsto \inf_{ v\in \rml^2([0,1],\msv):\eqsp g=\phi^v} \int_0^1 || v_\tau||_\msv\dd \tau
 \end{equation}
By deriving differential constraints related to the minimum of \eqref{eq:geodesics_original} and using Cauchy lipshcitz conditions, geodesics can be defined only by giving the starting point and the initial velocity $v_0\in \msv$ \cite{miller2006geodesic}, as straight lines in Euclidean space.
Denoting by $w(v_0)$ the geodesic starting from the identity with inital velocity $v_0$, the exponential map is defined as $\varphi^{\{v_0\}}\triangleq \phi^v$ and the previous matching problem becomes a \textit{geodesic shooting problem}:
 \begin{equation}
  \label{eq:geodesics_shooting}
  \inf_{v_0 \in \msv} \mathscr{L}(\varphi^{\{v_0\}}.\mathbf{x},\mathbf{y}).
 \end{equation}
 Using $\varphi^{\{v_0\}}$ instead of $\phi^v$ for any $v\in \rml^2([0,1],\msv)$ regularizes the problem and induces a sparse representation for the learning diffeomorphisms.
 Moreover, by setting $\msv$ as an RKHS, the geodesic shooting problem has a unique solution and becomes tractable, as described in the next section.

@@

\vspace{-1ex}
\paragraph{Discrete parametrization of diffeomorpshim.}
%\alain{adapt this without mentioning applicaiton to time seriess}
%\sam{Est ce qu'on devrait pas introduire la dicretization du graphe ici et dire que l'on peut utiliser le théoreme de représentation}
In this part, $\msv$ is chosen as an RKHS \cite{berlinet2011reproducing} generated by a smooth kernel $K$ (e.g., Gaussian). % $K(x,y)\triangleq \exp(-|x-y|^2/\sigma^2)\operatorname{Id}_{d'}$ for any $(x,y)\in(\Rset^{d'})^2$, where $\sigma^2>0$ is the kernel width and $\operatorname{Id}_{d'}$ is the identity matrix.
We follow \cite{durrleman2013sparse} and define a 
 discrete parameterization of the velocity fields to perform geodesics shooting \eqref{eq:geodesics_shooting}.  
  The initial velocity field $v_0$ is chosen as a finite linear combination of the RKHS basis vector fields, 
$\mathbf{n}_0$ control points $\msx_0=(x_{k,0})_{k\in[\mathbf{n}_0]}\in (\Rset^{d'})^{\mathbf{n}_0}$ and momentum vectors $\alpha_0=(\alpha_{k,0})_{k\in[\mathbf{n}_0]}\in (\Rset^{d'})^{\mathbf{n}_0} $ are defined such that for any $x\in \Rset^{d'}$, 
  \begin{equation}
    \label{eq:def_v0}
    v_0\left(\alpha_0,\msx_0\right)(x)=\sum_{k=1}^{\mathbf{n}_0} K(x,x_{k,0})\alpha_{k,0} \eqsp .
  \end{equation}
   In our applications, the control points $(x_{k,0})_{k\in[\mathbf{n}_0]}$ can be understood as the discretized graph $(t_k,\mathbf{s}_0(t_k))_{k\in[\mathbf{n}_0]}$ of a starting time series $\mathbf{s}_0$. 
  With this parametrization of $v_0$, \citet{miller2006geodesic} show that the velocity field $v$ of the solution of \eqref{eq:geodesics_shooting} keeps the same
  structure along time, such that for any $x\in\Rset^{d'}$ and $\tau\in[0,1]$, 
  \begin{equation}
    \label{eq:specific_form}
    v_\tau(x)=\sum_{k=1}^{\mathbf{n}_0} K(x,x_k(\tau))\alpha_{k}(\tau) \eqsp ,
  \end{equation}
  \begin{equation} 
    \label{eq:integration}
      \left\{
        \begin{aligned}
        & \frac{\dd x_k(\tau)}{\dd \tau}=v_\tau(x_k(\tau)) \eqsp, \quad
        \frac{\dd \alpha_k(\tau)}{\dd \tau}=-\sum_{k=1}^{\mathbf{n}_0} \dd_{x_k(\tau)} K(x_k(\tau),x_l(\tau))\alpha_{l}(\tau)^\top \alpha_{k}(\tau) \eqsp  \\
        & \alpha_k(0)=\alpha_{k,0},\quad x_k(0)=x_{k,0} \eqsp , k\in[\mathbf{n}_0] 
        \end{aligned}
        \right .
  \end{equation}
  These equations are derived from the hamiltonian $H:(\alpha_k,x_k)_{k\in [\mathbf{n}_0]}\mapsto \sum_{k,l=1}^{\mathbf{n}_0} \alpha_{k}^\top K(x_k,x_l)\alpha_{l}  $, such that
  the velocity norm is preserved $||v_\tau||_{\msv}=||v_0||_\msv $ for any $\tau\in [0,1]$.
   By \eqref{eq:integration}, the velocity field related to a geodesic $v^*$ is fully parametrized by its initial control points and momentum $(x_{k,0},\alpha_{k,0})_{k\in[\mathbf{n}_0]}$.
   Thus, given a set of targets $\mathbf{y}=(y_i)_{i\in[T_2]}$ in $\Rset^{d'}$, a set of starting points $\mathbf{x}=(x_{i,0})_{i\in[T_1]}$ in $\Rset^{d'}$, a RKHS's kernel $K:\Rset^{d'}\times \Rset^{d'}\to \Rset^{d'\times d'}$, a distance on sets $\mathscr{L}$, 
 a numerical integration scheme of ODE and a penalty factor $\lambda>0$, the basic geodesic shooting step minimizes the following function using a gradient descent method:
   \begin{equation}
    \label{eq:relaxation}
    \mathcal{F}_{\mathbf{x},\mathbf{y}}: (\alpha_k)_{k\in [T_1]}\mapsto \mathscr{L}\left(\varphi^{\{v_0\}}.\mathbf{x},\mathbf{y}\right)+\lambda||v_0||_\msv^2 \eqsp,  
   \end{equation}
   where $v_0$ is defined by \eqref{eq:def_v0} and $\varphi^{\{v_0\}}.\mathbf{x}$ is the result of the numerical integration of \eqref{eq:integration} using control points $\mathbf{x}$ and initial momentums $(\alpha_k)_{k\in[T_1]} $.

@@

 Note that \textbf{the number of samples $n_j$ is not necessary the same accross individuals} and the timestamps can be \textbf{irregularly sampled}.
%To ease the presentation of our methodology, we suppose that we have access to the continuous time series $(s^j)_{j\in[N]}$, while in practice, we can only access a $n_j$-samples $\tsig^j=(\tsig_i^j=s^j(t^j_i))_{i\in[n_j]}$ collected at timestamps $(t^j_i)_{i\in[n_j]}$ for any $j \in [N]$.
 %However, this difference does not change the overall rationale of our method and its implementation up to minor changes, which we summarize in \eqref{eq:general_optimization_problem}.

@@

     %\sam{Parler de l'unicté?}
     \vspace{-1ex}
     \paragraph{Optimization related to \eqref{eq:transformation}.}
     Defining the \textit{discretized graphs} of the time series $(s^j)_{j\in[N]}$ and a discretization of the reference graph $\msg(\mathbf{s}_0)$  as, for any $j\in[N]$,
     \begin{equation}
      \label{eq:descretized_graph}
      \mathbf{y}_j=\msg(\tsig^j)=(t_i^j,\tsig^j_i)_{ i\in[n_j]}\in (\Rset^{d+1})^{n_j},\quad \tilde{\msg}_0=(t_i^0,\tsig^0_i)_{i\in[\mathbf{n}_0]}\in (\Rset^{d+1})^{\mathbf{n}_0} \eqsp ,
     \end{equation}
       with $\mathbf{n}_0=\operatorname{median}((n_j)_{j\in[N]})$, the representation problem given in \eqref{eq:transformation} boils down solving:
      \begin{equation}
        \label{eq:general_optimization_problem}
        \argmin_{\tilde{\msg}_0,(\alpha_k^j)_{k\in [\mathbf{n}_0]}^{j\in[N]}} \sum_{j=1}^N \mathcal{F}_{\tilde{\msg}_0,\mathbf{y}_j}\left((\alpha_k^j)_{k\in [\mathbf{n}_0]}\right)\eqsp ,
      \end{equation}
      which is carried out by a gradient descent on the control points $\tilde{\msg}_0$ and the momentums $\mathbf{\alpha}_j=(\alpha_k^j)_{k\in [\mathbf{n}_0]}$ for any $j\in[N]$, initialized by a dataset's time series graph of size $\mathbf{n}_0$ and by $0_{(d+1)\mathbf{n}_0}$ respectively.
      The optimization hyperparameter details are given in \Cref{appendix:optimizers_details}.
     The result of the minimization $\tilde{\msg}_0$ is then considered as the $\mathbf{n}_0$-samples of a common time series $\mathbf{s}_0$ and the momentums $\mathbf{\alpha}_j$ encoding $\phi_j$ yields a feature vector in $\Rset^{d \mathbf{n}_0} $ of $s^j$ for any $j\in[N]$.
     %, which will be the representation of the time series when choosing the time series' graph sets as inputs, is dependent on the choice of kernel $K$ and the set distance $L$.
     Finally, the vectors $(\mathbf{\alpha}_j)_{j\in[N]}$ can be analyzed with any statistical or machine learning tools such as Principal Components Analysis (PCA), Latent Discriminant Analysis (LDA), longitudinal data analysis and so on.
%As $\bfs_0$ is supposed to be fixed, then the representation of the time series $(s_j)_{j\in[N]}$ boils down to the one of the transformation $(\phi_j)_{j\in[N]}$.
 %In \Cref{section:optimization} the choice of the common time series $\mathbf{s}_0$ will be optimized to have $\phi_j$ of minimal norm. %T is given in the following part.

@@

Nevertheless, \eqref{eq:general_optimization_problem} ask to define a kernel and a loss in order to perform geodesic shooting \ref{eq:relaxation}, which is the purpose of the next subsection.
\vspace{-1ex}
   \subsection{Application of LDDMM to time series analysis: TS-LDDMM}
   \vspace{-1ex}
%\tcr{Mettre l'applicaiton clair pour les times series}

        \label{section:time_series_specificity}
        In this section, we present our theoretical contribution: we tailor the LDDMM framework to handle time series data.
          The reason is that applying a general diffeomorphism $\phi$ from $\Rset^{d+1}$ to a time series' graph $\msg(s)$ can result in a set $\phi.\msg(s)$ that does not correspond to the graph of any time series, as illustrated in the \Cref{fig:diffeo}.
           Thus, Time series graph have more structure than a simple 1D curve \cite{glaunes2008large} and deserve their special analysis which will prove fruitful as demonstrated in \ref{section:experiments}.% thus this case is not an application of \cite{glaunes2004diffeomorphic}

@@

    %     Let $\bar{f}\in \rmC^0(\Rset^{d+1},\Rset^d)$, $\bar{\gamma}\in  \mcd(\Rset) $ and $\mathbf{s}_0\in \rmC^0(\msi,\Rset^d)$ with $\msi$ an interval of $\Rset$.
    %     We have :
    %     \begin{align}
    %       \Pi_{\gamma,f}.\msg(\mathbf{s}_0)&=\{(\gamma(t),f(t,\mathbf{s}_0(t))),\eqsp t\in \msi \} \\
    %       &\label{eq:proof1_last_eq}=\{(t,f\left(\gamma^{-1}(t),\mathbf{s}_0(\gamma^{-1}(t))\right),\eqsp t\in \gamma(\msi) \} \eqsp .
    %     \end{align}
    %     By defining $\bar{s}:t\in \gamma(\msi)\to f\left(\gamma^{-1}(t),\mathbf{s}_0(\gamma^{-1}(t))\right) $, we have $\bar{s}\in \rmC^0(\gamma(\msi), \Rset^d) $ by composition of continuous functions
    %     and $ \msg(\bar{s})= \Pi_{\gamma,f}.\msg(\mathbf{s}_0)$ by \eqref{eq:proof1_last_eq}, which concludes the proof.
    % \end{proof}
    \begin{remark}
      that for any $\gamma \in \mcd(\Rset) $ and $s\in \rmC^0(\msi,\Rset^d)$,
   \begin{equation}
    \{(\gamma(t),s(t)),\eqsp t\in \msi \}=\{(t,s\circ \gamma^{-1}(t)):\eqsp t\in\gamma(\msi) \}\eqsp .
   \end{equation}
   As a result, $\Psi_\gamma $ can be understood as a temporal reparametrization and $\Phi_f$ encodes the transformation about the space.
 \end{remark}

 \vspace{-1ex}
    \paragraph{Choice for the kernel associated with the RKHS $\msv$}
    \label{paragraph:kernel_V}
    As depicted on \Cref{fig:diffeo}-\ref{fig:transport}, we can not use any kernel $K$ to apply the previous methodology to learn deformations on time series' graphs.
    We describe and motivate our choice in this paragraph.
     Denote the one-dimensional Gaussian kernel by $K_\sigma^{(a)}(x,y)=\exp(-|x-y|^2/\sigma)$ for any $(x,y)\in (\Rset^a)^2$, $a\in \Nset$ and $\sigma>0$.
    To solve the geodesic shooting problem \eqref{eq:relaxation} on $\Rset^{d+1}$, we consider for $\msv$ the RKHS associated with the kernel defined for any $(t,x),(t',x')\in (\Rset^{d+1})^2$:
    \begin{align}
      \label{eq:kernel_TAS}
      &K_{\msg}((t,x),(t',x'))=\begin{pmatrix}
        c_0K_{\text{time}} & 0 \\
        0 & c_1 K_{\text{space}} 
        \end{pmatrix} \eqsp ,\\
       & K_{\text{space}}=K_{\sigma_{T,1}}^{(1)}(t,t')K_{\sigma_x}^{(d)}(x,x') \Idd\eqsp,K_{\text{time}}=K_{\sigma_{T,0}}^{(1)}(t,t') \eqsp,
    \end{align}
    parametrized by the widths $\sigma_{T,0},\sigma_{T,1},\sigma_x>0$ and the constants $c_0,c_1>0$.
This choice for $K_\msg$ is motivated by the representation \Cref{theorem:representation} and the following result. 
    \begin{lemma}
      \label{lemma:choice_of_kernel_V}
      If we denote by $\msv$ the RKHS associated with the kernel $K_{\msg}$, then for any vector field $v$ generated by \eqref{eq:integration} with $v_0$ satisfying \eqref{eq:def_v0},
       there exist $\gamma \in \msd(\Rset) $ and $f\in \rmC^1(\Rset^{d+1},\Rset^d)$ such that $\phi^v=\Psi_\gamma\circ\Phi_f $.
    \end{lemma}
    Instead of Gaussian kernels, other types of smooth kernel can be selected as long as the structure \eqref{eq:kernel_TAS} is respected.
    % \begin{proof}
    %   Let $v$ be a vector field generated by \eqref{eq:integration} with $v_0$ satisfying \eqref{eq:def_v0}.
    %  We remark that the first coordinate of the velocity field $v_\tau$ denoted by $v_\tau^{\text{time}}$ only depends on the time variable $t$ for any $\tau\in[0,1]$.
    %  Thus, when computing the first coordinate of the deformation $\phi^v$, denoted by $\gamma$, we integrate \eqref{eq:LDDMM_dynamic} with $v_\tau$ replaced by $v_\tau^{\text{time}}$,
    %   thus $\gamma$ is independant of the variable $x$. Moreover, $\gamma\in \mcd(\Rset)$ since a Gaussian kernel induced an Hilbert space $\msv$ satisfying $|f|_V\leq |f|_\infty+ |\dd f|_\infty  $ for any $f\in \msv$ by \citep[Theorem 9]{glaunes2005transport}.
    %   For the same reason, we have $\phi^v\in \mcd(\Rset^{d+1})$, and thus its last coordinates denoted by $f$ belongs to $\rmC^1(\Rset^{d+1},\Rset^d)$, and by construction $\phi^v=\Psi_\gamma\circ\Phi_f $.
    % \end{proof
    \begin{remark}
      With this choice of kernel, the features associated to the time transformation can be extracted from the momentums $(\alpha_{k,0})_{k\in[\mathbf{n}_0]}\in (\Rset^{d+1})^{\mathbf{n}_0}$ in \eqref{eq:def_v0} by taking the coordinates related to time.
      However, the features related to the space transformation are not only in the space coordinates since the related kernel $K_{\text{space}}$ depends on time as well.
     \end{remark}
    In \Cref{appendix:kernel_TS_LDDMM}, we give guidelines for selecting the hyperparameters $(\sigma_{T,0},\sigma_{T,1},\sigma_x,c_0,c_1)$.

@@

       Leaving the LDDMM representation, the results of \cite{srivastava2010shape,heo2024logistic} addresses the representation of curves with the Square-Root Velocity (SRV) representation.
        However, the SRV representation is applied after a reparametrization of the temporal dimension on the unit length segment.
         Consequently, the graph structure of the time series is not respected, and the original time evolution of the time series is not encoded in the final representation.
          Very recently, in a functional data analysis framework, a paper \cite{wu2024shape} (Shape-FPCA) improved by representing the original time evolution.
           Nevertheless, this method is made for \textit{continuous objects} and only applies to time series of \textit{same length}, making the estimation more sensitive to noise.

>>>>

 In the next section, time series are going to be represented by diffeomorphism parameters $(\alpha_j)_{j\in[N]}$.
 That is why LDDMM is chosen since it offers a parametrization for diffeomorphisms that is sparse and interpretable, two features particularly relevant in the biomedical context.

The basic problem that we consider in this section is the following. Given a set of targets $\mathbf{y}=(y_i)_{i\in[T_2]}$ in $\Rset^{d'}$\footnote{Note that we denote by $d'\in\nset$ the ambient space}, a set of starting points $\mathbf{x}=(x_{i})_{i\in[T_1]}$ in $\Rset^{d'}$, we aim to find a diffeomorphism $\phi$ such that the finite set of points $\mathbf{y}$ is similar in a certain sense to the set of finite sets of transformed points $\phi \cdot \mathbf{x} =(\phi(x_i))_{i\in[T_1]} $.
 The function $\phi$ is occasionally referred to as a \textit{deformation}. In general, these sets $\mathbf{x},\mathbf{y}$ are meshes of continuous objects, e.g., surfaces, curves, images, and so on.

@@

%\paragraph*{Regularizing the problem to derive uniqueness.}
It has been proposed in \cite{miller2006geodesic} to interpret $\msv$ as a tangent space relative to the group of diffeomorphisms $\msh=\{ \phi^v:\eqsp v\in \rml^2([0,1],\msv)\}$.
Following this geometric point of view, geodesics can be constructed on $\msh$ by using the following squared norm 
 \begin{equation}
  \label{eq:geodesics_original}
    \mathscr{R}^2: g\in \msh\mapsto \inf_{ v\in \rml^2([0,1],\msv):\eqsp g=\phi^v} \int_0^1 || v_\tau||_\msv\dd \tau
 \end{equation}
By deriving differential constraints related to the minimum of \eqref{eq:geodesics_original} and using Cauchy-Lipschitz conditions, geodesics can be defined only by giving the starting point and the initial velocity $v_0\in \msv$ \cite{miller2006geodesic}, as straight lines in Euclidean space.
Denoting by $w(v_0)$ the geodesic starting from the identity with inital velocity $v_0$, the exponential map is defined as $\varphi^{\{v_0\}}\triangleq \phi^v$ and the previous matching problem becomes a \textit{geodesic shooting problem}:
 \begin{equation}
  \label{eq:geodesics_shooting}
  \inf_{v_0 \in \msv} \mathscr{L}(\varphi^{\{v_0\}}.\mathbf{x},\mathbf{y}).
 \end{equation}
 Using $\varphi^{\{v_0\}}$ instead of $\phi^v$ for any $v\in \rml^2([0,1],\msv)$ regularizes the problem and induces a sparse representation for the learning diffeomorphisms.
 Moreover, by setting $\msv$ as an RKHS, the geodesic shooting problem has a unique solution and becomes tractable, as described in the next section.

@@

\vspace{-1ex}
\paragraph{Discrete parametrization of diffeomorpshim.}

In this part, $\msv$ is chosen as an RKHS \cite{berlinet2011reproducing} generated by a smooth kernel $K$ (e.g., Gaussian). 
We follow \cite{durrleman2013sparse} and define a 
 discrete parameterization of the velocity fields to perform geodesics shooting \eqref{eq:geodesics_shooting}.  
  The initial velocity field $v_0$ is chosen as a finite linear combination of the RKHS basis vector fields, 
$\mathbf{n}_0$ control points $\msx_0=(x_{k,0})_{k\in[\mathbf{n}_0]}\in (\Rset^{d'})^{\mathbf{n}_0}$ and momentum vectors $\alpha_0=(\alpha_{k,0})_{k\in[\mathbf{n}_0]}\in (\Rset^{d'})^{\mathbf{n}_0} $ are defined such that for any $x\in \Rset^{d'}$, 
  \begin{equation}
    \label{eq:def_v0}
    v_0\left(\alpha_0,\msx_0\right)(x)=\sum_{k=1}^{\mathbf{n}_0} K(x,x_{k,0})\alpha_{k,0} \eqsp .
  \end{equation}
   In our applications, the control points $(x_{k,0})_{k\in[\mathbf{n}_0]}$ can be understood as the discretized graph $(t_k,\mathbf{s}_0(t_k))_{k\in[\mathbf{n}_0]}$ of a starting time series $\mathbf{s}_0$. 
  With this parametrization of $v_0$, \citet{miller2006geodesic} show that the velocity field $v$ of the solution of \eqref{eq:geodesics_shooting} keeps the same
  structure along time, such that for any $x\in\Rset^{d'}$ and $\tau\in[0,1]$, 
  \begin{equation}
    \label{eq:specific_form}
    v_\tau(x)=\sum_{k=1}^{\mathbf{n}_0} K(x,x_k(\tau))\alpha_{k}(\tau) \eqsp ,
  \end{equation}
  \begin{equation} 
    \label{eq:integration}
      \left\{
        \begin{aligned}
        & \frac{\dd x_k(\tau)}{\dd \tau}=v_\tau(x_k(\tau)) \eqsp, \quad
        \frac{\dd \alpha_k(\tau)}{\dd \tau}=-\sum_{k=1}^{\mathbf{n}_0} \dd_{x_k(\tau)} K(x_k(\tau),x_l(\tau))\alpha_{l}(\tau)^\top \alpha_{k}(\tau) \eqsp  \\
        & \alpha_k(0)=\alpha_{k,0},\quad x_k(0)=x_{k,0} \eqsp , k\in[\mathbf{n}_0] 
        \end{aligned}
        \right .
  \end{equation}
  These equations are derived from the hamiltonian $H:(\alpha_k,x_k)_{k\in [\mathbf{n}_0]}\mapsto \sum_{k,l=1}^{\mathbf{n}_0} \alpha_{k}^\top K(x_k,x_l)\alpha_{l}  $, such that
  the velocity norm is preserved $||v_\tau||_{\msv}=||v_0||_\msv $ for any $\tau\in [0,1]$.
   By \eqref{eq:integration}, the velocity field related to a geodesic $v^*$ is fully parametrized by its initial control points and momentum $(x_{k,0},\alpha_{k,0})_{k\in[\mathbf{n}_0]}$.
   Thus, given a set of targets $\mathbf{y}=(y_i)_{i\in[T_2]}$ in $\Rset^{d'}$, a set of starting points $\mathbf{x}=(x_{i,0})_{i\in[T_1]}$ in $\Rset^{d'}$, a RKHS's kernel $K:\Rset^{d'}\times \Rset^{d'}\to \Rset^{d'\times d'}$, a distance on sets $\mathscr{L}$, 
 a numerical integration scheme of ODE and a penalty factor $\lambda>0$, the basic geodesic shooting step minimizes the following function using a gradient descent method:
   \begin{equation}
    \label{eq:relaxation}
    \mathcal{F}_{\mathbf{x},\mathbf{y}}: (\alpha_k)_{k\in [T_1]}\mapsto \mathscr{L}\left(\varphi^{\{v_0\}}.\mathbf{x},\mathbf{y}\right)+\lambda||v_0||_\msv^2 \eqsp,  
   \end{equation}
   where $v_0$ is defined by \eqref{eq:def_v0} and $\varphi^{\{v_0\}}.\mathbf{x}$ is the result of the numerical integration of \eqref{eq:integration} using control points $\mathbf{x}$ and initial momentums $(\alpha_k)_{k\in[T_1]} $.

@@

 Note that \textbf{the number of samples $n_j$ is not necessarily the same across individuals} and the timestamps can be \textbf{irregularly sampled}.

@@

     %\sam{Parler de l'unicté?}
     \vspace{-1ex}
     \paragraph{Optimization related to \eqref{eq:transformation}.}
     Defining the \textit{discretized graphs} of the time series $(s^j)_{j\in[N]}$ and a discretization of the reference graph $\msg(\mathbf{s}_0)$ as, for any $j\in[N]$,
 \begin{equation}
 \label{eq:descretized_graph}
 \mathbf{y}_j=\msg(\tsig^j)=(t_i^j,\tsig^j_i)_{ i\in[n_j]}\in (\Rset^{d+1})^{n_j},\quad \tilde{\msg}_0=(t_i^0,\tsig^0_i)_{i\in[\mathbf{n}_0]}\in (\Rset^{d+1})^{\mathbf{n}_0} \eqsp ,
 \end{equation}
 with $\mathbf{n}_0=\operatorname{median}((n_j)_{j\in[N]})$, the representation problem given in \eqref{eq:transformation} boils down solving:
 \begin{equation}
 \label{eq:general_optimization_problem}
 \argmin_{\tilde{\msg}_0,(\alpha_k^j)_{k\in [\mathbf{n}_0]}^{j\in[N]}} \sum_{j=1}^N \mathcal{F}_{\tilde{\msg}_0,\mathbf{y}_j}\left((\alpha_k^j)_{k\in [\mathbf{n}_0]}\right)\eqsp ,
 \end{equation}
 which is carried out by gradient descent on the control points $\tilde{\msg}_0$ and the momentums $\mathbf{\alpha}_j=(\alpha_k^j)_{k\in [\mathbf{n}_0]}$ for any $j\in[N]$, initialized by a dataset's time series graph of size $\mathbf{n}_0$ and by $0_{(d+1)\mathbf{n}_0}$ respectively.
 The optimization hyperparameter details are given in \Cref{appendix:optimizers_details}.
 The result of the minimization $\tilde{\msg}_0$ is then considered as the $\mathbf{n}_0$-samples of a common time series $\mathbf{s}_0$ and the momentums $\mathbf{\alpha}_j$ encoding $\phi_j$ yields a feature vector in $\Rset^{d \mathbf{n}_0} $ of $s^j$ for any $j\in[N]$.
 Finally, the vectors $(\mathbf{\alpha}_j)_{j\in[N]}$ can be analyzed with any statistical or machine learning tools such as Principal Components Analysis (PCA), Latent Discriminant Analysis (LDA), longitudinal data analysis and so on.
%As $\bfs_0$ is supposed to be fixed, then the representation of the time series $(s_j)_{j\in[N]}$ boils down to the one of the transformation $(\phi_j)_{j\in[N]}$.
 %In \Cref{section:optimization} the choice of the common time series $\mathbf{s}_0$ will be optimized to have $\phi_j$ of minimal norm. %T is given in the following part.

@@

Nevertheless, \eqref{eq:general_optimization_problem} asks to define a kernel and a loss in order to perform geodesic shooting \ref{eq:relaxation}, which is the purpose of the following subsection.
\vspace{-1ex}
   \subsection{Application of LDDMM to time series analysis: TS-LDDMM}
   \vspace{-1ex}
%\tcr{Mettre l'applicaiton clair pour les times series}

        \label{section:time_series_specificity}
        This section presents our theoretical contribution: we tailor the LDDMM framework to handle time series data.
        The reason is that applying a general diffeomorphism $\phi$ from $\Rset^{d+1}$ to a time series' graph $\msg(s)$ can result in a set $\phi.\msg(s)$ that does not correspond to the graph of any time series, as illustrated in the \Cref{fig:diffeo}.
        Thus, time series graphs have more structure than a simple 1D curve \cite{glaunes2008large} and deserve their unique analysis, which will prove fruitful as demonstrated in \ref{section:experiments}.

@@

    %     Let $\bar{f}\in \rmC^0(\Rset^{d+1},\Rset^d)$, $\bar{\gamma}\in  \mcd(\Rset) $ and $\mathbf{s}_0\in \rmC^0(\msi,\Rset^d)$ with $\msi$ an interval of $\Rset$.
    %     We have :
    %     \begin{align}
    %       \Pi_{\gamma,f}.\msg(\mathbf{s}_0)&=\{(\gamma(t),f(t,\mathbf{s}_0(t))),\eqsp t\in \msi \} \\
    %       &\label{eq:proof1_last_eq}=\{(t,f\left(\gamma^{-1}(t),\mathbf{s}_0(\gamma^{-1}(t))\right),\eqsp t\in \gamma(\msi) \} \eqsp .
    %     \end{align}
    %     By defining $\bar{s}:t\in \gamma(\msi)\to f\left(\gamma^{-1}(t),\mathbf{s}_0(\gamma^{-1}(t))\right) $, we have $\bar{s}\in \rmC^0(\gamma(\msi), \Rset^d) $ by composition of continuous functions
    %     and $ \msg(\bar{s})= \Pi_{\gamma,f}.\msg(\mathbf{s}_0)$ by \eqref{eq:proof1_last_eq}, which concludes the proof.
    % \end{proof}
    \begin{remark}
      Note that for any $\gamma \in \mcd(\Rset) $ and $s\in \rmC^0(\msi,\Rset^d)$,
   \begin{equation}
    \{(\gamma(t),s(t)),\eqsp t\in \msi \}=\{(t,s\circ \gamma^{-1}(t)):\eqsp t\in\gamma(\msi) \}\eqsp .
   \end{equation}
   As a result, $\Psi_\gamma $ can be understood as a temporal reparametrization and $\Phi_f$ encodes the transformation about the space.
 \end{remark}

 \vspace{-1ex}
    \paragraph{Choice for the kernel associated with the RKHS $\msv$}
    \label{paragraph:kernel_V}
    As depicted on \Cref{fig:diffeo}-\ref{fig:transport}, we can not use any kernel $K$ to apply the previous methodology to learn deformations on time series' graphs.
    We describe and motivate our choice in this paragraph.
     Denote the one-dimensional Gaussian kernel by $K_\sigma^{(a)}(x,y)=\exp(-|x-y|^2/\sigma)$ for any $(x,y)\in (\Rset^a)^2$, $a\in \Nset$ and $\sigma>0$.
    To solve the geodesic shooting problem \eqref{eq:relaxation} on $\Rset^{d+1}$, we consider for $\msv$ the RKHS associated with the kernel defined for any $(t,x),(t',x')\in (\Rset^{d+1})^2$:
    \begin{align}
      \label{eq:kernel_TAS}
      &K_{\msg}((t,x),(t',x'))=\begin{pmatrix}
        c_0K_{\text{time}} & 0 \\
        0 & c_1 K_{\text{space}} 
        \end{pmatrix} \eqsp ,\\
       & K_{\text{space}}=K_{\sigma_{T,1}}^{(1)}(t,t')K_{\sigma_x}^{(d)}(x,x') \Idd\eqsp,K_{\text{time}}=K_{\sigma_{T,0}}^{(1)}(t,t') \eqsp,
    \end{align}
    parametrized by the widths $\sigma_{T,0},\sigma_{T,1},\sigma_x>0$ and the constants $c_0,c_1>0$.
This choice for $K_\msg$ is motivated by the representation \Cref{theorem:representation} and the following result. 
    \begin{lemma}
      \label{lemma:choice_of_kernel_V}
      If we denote by $\msv$ the RKHS associated with the kernel $K_{\msg}$, then for any vector field $v$ generated by \eqref{eq:integration} with $v_0$ satisfying \eqref{eq:def_v0},
       there exist $\gamma \in \msd(\Rset) $ and $f\in \rmC^1(\Rset^{d+1},\Rset^d)$ such that $\phi^v=\Psi_\gamma\circ\Phi_f $.
    \end{lemma}
    Instead of Gaussian kernels, other types of smooth kernels can be selected as long as the structure \eqref{eq:kernel_TAS} is respected.
    % \begin{proof}
    %   Let $v$ be a vector field generated by \eqref{eq:integration} with $v_0$ satisfying \eqref{eq:def_v0}.
    %  We remark that the first coordinate of the velocity field $v_\tau$ denoted by $v_\tau^{\text{time}}$ only depends on the time variable $t$ for any $\tau\in[0,1]$.
    %  Thus, when computing the first coordinate of the deformation $\phi^v$, denoted by $\gamma$, we integrate \eqref{eq:LDDMM_dynamic} with $v_\tau$ replaced by $v_\tau^{\text{time}}$,
    %   thus $\gamma$ is independant of the variable $x$. Moreover, $\gamma\in \mcd(\Rset)$ since a Gaussian kernel induced an Hilbert space $\msv$ satisfying $|f|_V\leq |f|_\infty+ |\dd f|_\infty  $ for any $f\in \msv$ by \citep[Theorem 9]{glaunes2005transport}.
    %   For the same reason, we have $\phi^v\in \mcd(\Rset^{d+1})$, and thus its last coordinates denoted by $f$ belongs to $\rmC^1(\Rset^{d+1},\Rset^d)$, and by construction $\phi^v=\Psi_\gamma\circ\Phi_f $.
    % \end{proof
    \begin{remark}
      With this choice of kernel, the features associated with the time transformation can be extracted from the momentums $(\alpha_{k,0})_{k\in[\mathbf{n}_0]}\in (\Rset^{d+1})^{\mathbf{n}_0}$ in \eqref{eq:def_v0} by taking the coordinates related to time.
      However, the features related to the space transformation are not only in the space coordinates since the related kernel $K_{\text{space}}$ depends on time as well.
      \end{remark}
      In \Cref{appendix:kernel_TS_LDDMM}, we give guidelines for selecting the hyperparameters $(\sigma_{T,0},\sigma_{T,1},\sigma_x,c_0,c_1)$.

@@

 Leaving the LDDMM representation, the results of \cite{srivastava2010shape,heo2024logistic} address the representation of curves with the Square-Root Velocity (SRV) representation.
 However, the SRV representation is applied after reparametrization of the temporal dimension of the unit length segment.
 Consequently, the graph structure of the time series is not respected, and the original time evolution of the time series is not encoded in the final representation.
 Very recently, in a functional data analysis framework, a paper \cite{wu2024shape} (Shape-FPCA) improved by representing the original time evolution.
 Nevertheless, this method is made for \textit{continuous objects} and only applies to time series of \textit{same length}, making the estimation more sensitive to noise.
